---
title: "Data Visualization - Mini-Project 2"
author: "Nico Seddio `nseddio0283@floridapoly.edu`"
output:
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

## Setup

```{r setup, message=FALSE}
library(tidyverse)
# install.packages("sf")
library(sf)
# install.packages("plotly")
library(plotly)
lakes <- st_read("../data/Florida_Lakes/Florida_Lakes.shp")
counties <- st_read("../data/Florida_Counties/Florida_Counties.shp")
schools <- st_read("../data/Public_Private_Schools_in_Florida-2017/gc_schools_sep17.shp")
# Some functions on lakes return an error about a bad vertex:
# Error in wk_handle.wk_wkb(wkb, s2_geography_writer(oriented = oriented, :
# Loop 0 is not valid: Edge 382 has duplicate vertex with edge 385
# https://stackoverflow.com/questions/79756123/what-issue-is-st-make-valid-fixing-when-applied-to-osm-data
lakes <- st_make_valid(lakes)
```

## Summarize

Across the three datasets, the `COUNTY` name is our best join key. The `Florida_Counties` dataset calls this `COUNTYNAME`, and counties `DADE`, `ST. JOHNS`, and `ST. LUCIE` are mismatched in spelling between the datasets. We fix these issues before working with the data and do a quick check to make sure no other mismatches in county name exist (an empty result is good).

```{r counties_fix}
# Rename COUNTY -> COUNTY_ID and COUNTYNAME -> COUNTY so COUNTY is consistent across data
counties <- counties %>% 
  mutate(COUNTY_ID = COUNTY, COUNTY = COUNTYNAME) %>% 
  select(-COUNTYNAME)
# Fix mismatched spellings to match Florida Counties source of truth
schools <- schools %>% 
  mutate(COUNTY = recode_values(COUNTY,
    "ST JOHNS" ~ "ST. JOHNS",
    "ST LUCIE" ~ "ST. LUCIE",
    "MIAMI-DADE" ~ "DADE",
    default = COUNTY))
lakes <- lakes %>% 
  mutate(COUNTY = recode_values(COUNTY,
    "ST. JOHN" ~ "ST. JOHNS",
    "MIAMI-DADE" ~ "DADE",
    default = COUNTY))
# Review for mismatches, empty is good
counties %>% 
  distinct(county = COUNTY) %>% 
  arrange(county) %>% 
  full_join(
    schools %>% distinct(school_county = COUNTY),
    by = join_by(county == school_county),
    keep = TRUE) %>% 
  full_join(
    lakes %>% distinct(lake_county = COUNTY),
    by = join_by(county == lake_county),
    keep = TRUE) %>% 
  filter(is.na(county) | is.na(school_county) | is.na(lake_county))
```

#### Group lakes into counties
```{r lakes_by_county, message=FALSE}
lakes_by_county <- lakes %>% 
  group_by(COUNTY) %>% 
  summarise(lake_count = n(),
            total_lake_area = sum(SHAPEAREA))
lakes_by_county %>% head(5)
```

#### Group schools into counties

```{r schools_by_county, message=FALSE}
schools_by_county <- schools %>% 
  group_by(COUNTY) %>% 
  summarise(school_count = n(),
            total_enrollment = sum(ENROLLMENT, na.rm = TRUE))
schools_by_county %>% head(5)
```

## Visualizations

#### Total enrollment in county with school count

The horizontal bar chart below shows the top 15 counties by total enrollment, 
with Dade county at the top with the most number of schools and the most number 
of enrollments. The interactivity allows users to review raw numbers of both 
enrollment and school count without these values cluttering the chart.

```{r bar_fifteen, fig.alt="Interactive horizontal bar chart showing top 15 counties by total enrollment with Dade county at the top with the most number of schools and the most number of enrollments."}
enrollment_bar <- schools_by_county %>% 
  slice_max(total_enrollment, n=15) %>% 
  ggplot(aes(
    total_enrollment,
    fct_reorder(COUNTY, total_enrollment), 
    fill = school_count)) +
  geom_col() +
  scale_fill_viridis_c() +
  labs(
    title = "County Enrollment Size and School Count", 
    fill = "Schools", 
    x = "Total Enrollment", 
    y = "County") + 
  scale_x_continuous(labels = scales::comma) +
  theme_minimal()
#enrollment_bar
ggplotly(enrollment_bar)
#save self-contained
htmlwidgets::saveWidget(ggplotly(enrollment_bar),"interactive-enrollment-bar.html", selfcontained = TRUE)
```

#### Five largest counties with operating enrollment

Column chart showing top 5 largest counties by total enrollment, each column as 
a county showing percentage division by school type (public, private, charter, 
etc.). The interactivity here allows presenting the data in a visually interesting
way as percentages of a whole while still allowing the user to drill down and get
actual values like total enrollment and school operating type.

```{r col_five, fig.alt="Column chart showing top 5 largest counties by total enrollment, each column as a county showing percentage division by school type (public, private, charter, etc.)."}
enrollment_col <- schools %>% 
  group_by(COUNTY,OPERATING) %>% 
  summarise(total_enrollment = sum(ENROLLMENT, na.rm = TRUE)) %>% 
  filter(total_enrollment != 0) %>% 
  distinct(COUNTY,OPERATING,total_enrollment) %>% 
  inner_join(
    schools_by_county %>% 
      slice_max(total_enrollment, n=5) %>% 
      distinct(COUNTY)) %>% 
  ggplot() +
  geom_col(aes(x = COUNTY, y = total_enrollment, fill = OPERATING), position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Largest Counties by Enrollment with Type", 
    x = "", 
    y = "Enrollment", fill = "School Type") +
  scale_fill_viridis_d() +
  theme_minimal()
#enrollment_col
ggplotly(enrollment_col)
htmlwidgets::saveWidget(ggplotly(enrollment_col),"interactive-enrollment-col.html", selfcontained = TRUE)
```


#### Lakes and Schools of Central Florida

Plotting all the counties of Florida together on one map would be difficult to 
look at. So let's plot just the data around Florida Polytechnic University.

```{r map_florida, fig.alt="Map of central Florida from Pinellas to Osceola, Volusia to Hardee, with distribution of schools and lakes. Florida Poly is pointed out towards the center of the map.",message=FALSE}
florida_poly <- st_sfc(st_point(c(-81.9498, 28.1499)), crs = st_crs(counties))
central_florida <- counties %>% ggplot() +
  geom_sf(color = "white") +
  geom_sf(data = lakes, fill = "lightblue", color = NA) +
  geom_sf(data = schools, color = "black", alpha = 0.1) +
  geom_sf_label(aes(label = COUNTY), size = 2.5, color = "blue", alpha = .3) +
  geom_sf(data = florida_poly, color = "red", size = 4, shape = 18) +
  coord_sf(xlim = c(-81,-83), ylim = c(27.5,29)) +
  labs(title = "Central Florida Lakes and Schools", x = "", y = "") +
  annotate("text", x = -81.9, y = 28.25, label = "Florida Poly", color = "red") +
  theme_minimal()
central_florida
ggsave("../figures/project-02-central-florida.png", width = 8, height = 5, dpi = 150)
```

Interactive version of the above chart. Adding interaction here is neat because 
it allows us to start the visualization centered on where we want users to look 
to tell the story, but it also allows them to zoom out and scroll around the data
to see how our story lines up with the rest of the data. Not all the features of
the original map are supported by default though, so this demonstrates the extra 
work required to make sure the graph is compatible with interactive tools.

```{r map_florida_interactive, fig.alt="Interactive version of the map of central Florida from Pinellas to Osceola, Volusia to Hardee, with distribution of schools and lakes. Florida Poly is pointed out towards the center of the map.",message=FALSE}
ggplotly(central_florida)
htmlwidgets::saveWidget(ggplotly(central_florida),"interactive-central-florida.html", selfcontained = TRUE)
```

#### Does lake area have an impact on total enrollment?

Plotting the relationship.

```{r scatter_enrollment, fig.alt="Scatter plot of county total lake area (x) against total school enrollment in thousands (y) with a linear trend line; the relationship is weak and pulled by a few high-enrollment counties such as Dade and Broward, which are labeled."}
county_model_data <- counties %>% 
  distinct(COUNTY) %>% 
  inner_join(
    lakes_by_county %>% 
      distinct(COUNTY,total_lake_area,lake_count),
    by = join_by(COUNTY)) %>% 
  inner_join(
    schools_by_county %>% 
      distinct(COUNTY,total_enrollment,school_count),
    by = join_by(COUNTY))
county_model_data %>% 
  ggplot(aes(total_lake_area/1e6,total_enrollment/1000)) +
  geom_point() + 
  geom_smooth(method="lm") +
  geom_text(
    aes(label = if_else(total_enrollment > 150000, COUNTY, "")), 
    nudge_y = 15, 
    size = 3) +
  labs(
    title = "County Lake Area by Total Enrollment", 
    x = "Total Lake Area (degrees² × 10⁻⁶)", 
    y = "Total Enrollment (thousands)") +
  theme_minimal()
```

Creating a model.

```{r coefficient_enrollment, fig.alt="Coefficient plot of the estimated effect of total lake area on enrollment with its 95% confidence interval, sitting just above the dashed zero line — a weak, marginally positive effect."}
model <- lm(
  total_enrollment ~ total_lake_area, 
  data = county_model_data)
broom::tidy(model, conf.int = TRUE) %>% 
  filter(term != "(Intercept)") %>% 
  ggplot() +
  geom_pointrange(aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "County Lake Area Effect on Enrollment") +
  theme_minimal()
```

Clustering the plot with k-means.

```{r kmeans_clustering, fig.alt="Grid of nine scatter plots (k = 1 to 9) coloring counties by k-means cluster on lake area versus enrollment. Miami-Dade splits off early and the rest stay tightly grouped, so higher k adds little structure."}
mini_county_model_data <- county_model_data %>% 
  select(total_lake_area,total_enrollment)
kclusts <- tibble(k = 1:9) %>% 
  mutate(
    kclust = map(k, ~kmeans(mini_county_model_data, .x)),
    tidied = map(kclust, broom::tidy),
    glanced = map(kclust, broom::glance),
    augmented = map(kclust, broom::augment, mini_county_model_data)
  )
kclusts %>% 
  unnest(cols = c(augmented)) %>% 
  ggplot(aes(total_lake_area/1e6,total_enrollment/1000)) +
  geom_point(aes(color = .cluster), alpha = .8) +
  facet_wrap(~ k) +
  geom_point(data = kclusts %>% unnest(cols = c(tidied)), size = 5, shape = "x") +
  labs(
    title = "County Clustering by Lake Area and Enrollment", 
    x = "Total Lake Area (degrees² × 10⁻⁶)", 
    y = "Total Enrollment (thousands)") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_viridis_d() +
  theme_minimal()
```


## Redesigning a Bad Chart (Before/After)

A common way to mislead with an otherwise honest dataset is a truncated axis. 
Here is the top-five-county enrollment shown badly, then corrected.

```{r redesign_data}
top5 <- schools_by_county %>%
  st_drop_geometry() %>%
  slice_max(total_enrollment, n = 5) %>%
  mutate(COUNTY = fct_reorder(COUNTY, total_enrollment))
```

**Before — truncated axis and distracting fill:**

```{r bad_chart, fig.alt="A column chart of the five highest-enrollment counties whose y-axis starts at 220,000 instead of zero. The truncation makes Broward and Hillsborough look like tiny fractions of Dade even though their enrollments are much closer than that. Each bar is a different color."}
top5 %>% 
  ggplot(aes(COUNTY, total_enrollment, fill = COUNTY)) +
  geom_col() +
  coord_cartesian(ylim = c(220000, 460000)) +   # truncated window
  scale_fill_viridis_d() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Enrollment by County", x = NULL, y = "Enrollment") +
  theme_minimal() +
  theme(legend.position = "none")
```

**What's wrong with it:** the y-axis starts at 220,000, so bar *heights* no 
longer encode the actual values. Broward looks like a sliver of Dade when its 
enrollment is actually more than half. The colors of the bars differentiate them, 
but also suggest a value hierarchy where there is none.

**After — zero baseline, single accessible hue:**

```{r good_chart, fig.alt="A column chart of the same five counties with the y-axis starting at zero and exact enrollment labeled above each bar. Now the bars are proportional: Dade is clearly the largest but Broward and Hillsborough are visibly comparable, not negligible. All bars share one color."}
top5 %>% 
  ggplot(aes(COUNTY, total_enrollment)) +
  geom_col(fill = "grey") +
  geom_text(
    aes(label = scales::comma(total_enrollment)), 
    vjust = -0.4, 
    size = 3) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Enrollment by County (Top 5)", x = NULL, y = "Enrollment") +
  theme_minimal()
```

**How the redesign fixes it:** anchoring the y-axis at zero restores honest, 
proportional bar heights; value labels make the numbers exact; and dropping the 
color fill for a single colorblind-safe grey removes redundant, distracting color 
and any perceived bias.

## Report

#### Data & Process

I joined three Florida spatial datasets — county boundaries, lakes, and
public/private schools — on county name. The join surfaced three name mismatches
(`ST JOHNS`/`ST LUCIE` missing periods, `MIAMI-DADE` vs. `DADE`), which I found by
`full_join`-ing all three and filtering for `NA` matches, then normalized with
`recode_values()`. One invalid lake geometry was repaired with `st_make_valid()`.
I started without a fixed question and let the lakes-vs-schools comparison drive it.

#### What the Charts Show

- **Enrollment.** Dade leads on both enrollment and school count by a wide margin,
  and among the top five counties, district-run public schools account for nearly
  all enrollment. The interactive bars let the reader hover for exact counts and
  the school-count fill — values a static chart would force them to estimate.
  
- **Map + model.** Schools and lakes both concentrate in the same central areas,
  but the linear model finds only a weak, barely-positive lake-area effect, and
  k-means isolates Dade as its own cluster while every other county stays tightly
  grouped. The likeliest explanation is a shared driver — population and county
  size — not a real lake-to-school link; coastal water, which the data ignores,
  probably matters more than inland lakes. I read the map pattern as a population
  proxy, not evidence of a relationship.

#### Redesign & Accessibility

The before/after redesign takes the top-five enrollment bars, shows them with a
truncated axis and distracting fill, then fixes both: a zero baseline restores
proportional heights and a single colorblind-safe hue removes the false hierarchy
that per-bar color implied. Accessibility runs through the whole report — every
figure uses a `viridis` palette and carries `fig.alt` text, and color is never the
only cue (the redesign labels values directly).

#### Design Decisions & Next Steps

Plots share `theme_minimal()` and use color purposefully: a continuous `viridis`
scale encodes school count in the enrollment bars, and the map orients the reader
through position and one labeled reference point (Florida Poly) rather than color.
With more time I'd add county population and land area to separate "lakes" from
"size," and measure school-to-water distance directly instead of aggregating to
the county.
